Future Cosmic Microwave Background delensing with galaxies surveys.

### 2017-05-11 Notes/DES_rec_phi_paper.tex
text

The measurement of the CMB polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. A particular component, the so called B-modes is indeed a direct signature of the presence of gravitational waves in the early universe. However improving the level of noise is not enough. This is even more true if the aim is to not only detect the amplitude of gravitational waves but also the shape of their spectrum to test for example inflation consistency relations.
Removing the lensing component from the measurement of CMB B-modes will be important to constraint the amplitude of the primordial gravitational wave contribution to the signal.
Here we discuss the role of current and future large scale structure surveys  in improving the reconstruction of the lensing potential that lenses the CMB photons.
We find that..
\end{abstract}

@@

The full B-mode power spectrum measured on the sky is given by a possible primordial component $C_l^{BB,r}$ together with the lensing $C_l^{BB,\mathrm{lens}}$ contribution and instrumental noise $N_l^{BB}$:
\be
C_l^{BB,\mathrm{full}} = C_l^{BB,r} + C_l^{BB,\mathrm{lens}} + N_l^{BB}.
\ee
Given the current constrains, the lensing component is a significant source of B-modes that, for GW searches purposes correspond to a white noise source of roughly $5 \mu K$-arcmin. This means that it is not only bigger than the biggest allowed GW contribution at scale bigger than several degrees but it is comparable with the decreasing level of instrumental noise. 
For this reason it is very important to characterize it and eventually remove it form the data.
This can be done for example using \refeq{blens} (or its real-space analogue) given a measurement of the E-mode field and the lensing potential $\phi$.
While E is measured directly, we ca estimate $\phi$ using "tracers" of the dark matter distribution that creates the potential.

@@

Note that gain we have in adding a new tracer is not only proportional to its correlation with the CMB lensing but it also depends on how much it is correlated with the already used set of tracers.
\reffig{kernel} show the different kernels as a function of redshift computed using the models and parameters described in the following sections.
Galaxies clustering surveys can only reconstruct the low-z portion of the lensing kernel as can be seen from the LSST DES and DESI curves.
However given the significant low noise of this measurement they can still help the delensing of the CMB.
Furthermore the low-redshift tail of the lensing kernel is also the one affected by large scale structures non-linearities. They can indeed be extremely helpful if future CMB surveys will try to remove that contribution instead of attempting to model it.
On the other hand CIB and 21 cm surveys are target to high redshift structures and independently from the model assumed they show a fairly good overlap with the CMB lensing kernel.

@@

Very recently delensing has been proven possible on real data using both the internal CMB lensing potential reconstruction as well as external tracers.
Here we discuss the improvement due to combining this two probes.
This is already possible both on full sky using Planck data or in smaller sky patches combining Herschel or Planck CIB data with SPT and BK data.
Furthermore we test how much current relative low redshift galaxies surveys like D.E.S can improve delensing efficiency on smaller patch of the sky like the Bicep-Keck SPT-Pol footprint.

We model the current generation pf CMB experiments after 3 experiments: a full sky Planck experiment together with a deep low noise experiemtn and a high resolution one on a smaller patch of the sky.

@@

\section{Conclusions}
Delensing or more in general the ability to separate the lensing component of the B-mode from a possible primordial inflationary signal is crucial to fully exploit the capabilities of future experiments.
In this paper we studied the possible impact of large scale structure surveys in this important endeavor.
For current experiment CIB data had already proven to be very efficient in delensing. If no High res experiment is avaiable to reconstruct the lensing potential phi even a low-z experiment like D.E.S will improve th delensign efficiency by $~10\%$ if combined with Planck lensing and a CIB tracers.
As expected the lower the noise in the CMB experiment the more the delensign efficiency will be predominately come from the internal CMB reconstruction of the lensing potential. Indeed using the CMB itself we reconstruct only the lenses that actually lens the CMB obtained an almost perfect cleaning in the absence of noise. This approach will need a careful study of possible biases coming from using the same source (the CMB) we are delensing to reconstruct the lensing effect itself. This has already been applied to data and studied. However the fairly good efficiency of galaxies tracers might come in end to cross-check these internal biases. This will probably be needed to confirm a possible detection of gravitation waves if this rely heavily on delensing.

%=================================================================
\section{Acknowledgments}
We thank 
This work was partially supported by the Kavli Institute for Cosmological Physics at the University of Chicago through grants NSF PHY-1125897 and an endowment from the Kavli Foundation and its founder Fred Kavli.
%%%=================================================================

>>>>

The measurement of the CMB polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. A particular component, the so-called B-modes is indeed a direct signature of the presence of gravitational waves in the early universe. However improving the level of noise is not enough. This is even truer if the aim is to not only detect the amplitude of gravitational waves but also the shape of their spectrum to test for example inflation consistency relations.
Removing the lensing component from the measurement of CMB B-modes will be important to constrain the amplitude of the primordial gravitational wave contribution to the signal.
Here we discuss the role of current and future large scale structure surveys in improving the reconstruction of the lensing potential that lenses the CMB photons.
We find that..
\end{abstract}

@@

The full B-mode power spectrum measured on the sky is given by a possible primordial component $C_l^{BB,r}$ together with the lensing $C_l^{BB,\mathrm{lens}}$ contribution and instrumental noise $N_l^{BB}$:
\be
C_l^{BB,\mathrm{full}} = C_l^{BB,r} + C_l^{BB,\mathrm{lens}} + N_l^{BB}.
\ee
Given the current constraints, the lensing component is a significant source of B-modes that, for GW searches purposes correspond to a white noise source of roughly $5 \mu K$-arcmin. This means that it is not only bigger than the biggest allowed GW contribution at scales bigger than several degrees but it is comparable with the decreasing level of instrumental noise. 
For this reason, it is very important to characterize it and eventually remove it from the data.
This can be done for example using \refeq{blens} (or its real-space analog) given a measurement of the E-mode field and the lensing potential $\phi$.
While E is measured directly, we ca estimate $\phi$ using "tracers" of the dark matter distribution that creates the potential.

@@

Note that gain we have in adding a new tracer is not only proportional to its correlation with the CMB lensing but it also depends on how much it is correlated with the already used set of tracers.
\reffig{kernel} show the different kernels as a function of redshift computed using the models and parameters described in the following sections.
Galaxies clustering surveys can only reconstruct the low-z portion of the lensing kernel as can be seen from the LSST DES and DESI curves.
However given the significant low noise of this measurement they can still help the delensing of the CMB.
Furthermore, the low-redshift tail of the lensing kernel is also the one affected by large scale structures non-linearities. They can indeed be extremely helpful if future CMB surveys will try to remove that contribution instead of attempting to model it.
On the other end, CIB and 21 cm surveys are probing to high redshift structures and independently from the model assumed they show a fairly good overlap with the CMB lensing kernel.

@@

Very recently delensing has been proven possible on real data using both the internal CMB lensing potential reconstruction as well as external tracers.
Here we discuss the improvement due to combining this two probes.
This is already possible both on full sky using Planck data or in smaller sky patches combining Herschel or Planck CIB data with SPT and BK data.
Furthermore, we test how much current relative low redshift galaxies surveys like D.E.S can improve delensing efficiency on a smaller patch of the sky like the Bicep-Keck SPT-Pol footprint.

We model the current generation of CMB experiments after 3 experiments: a full sky Planck experiment together with a deep low noise experiment and a high resolution one on a smaller patch of the sky.

@@

\section{Conclusions}
Delensing or more in general the ability to separate the lensing component of the B-mode from a possible primordial inflationary signal is crucial to fully exploit the capabilities of future experiments.
In this paper, we studied the possible impact of large scale structure surveys in this important endeavor.
For current experiment, CIB data had already proven to be very efficient in delensing. If no High res experiment is available to reconstruct the lensing potential phi even a low-z experiment like D.E.S will improve the delensign efficiency by $~10\%$ if combined with Planck lensing and a CIB tracers.
As expected the lower the noise in the CMB experiment the more the delensign efficiency will be predominately come from the internal CMB reconstruction of the lensing potential. Indeed using the CMB itself we reconstruct only the lenses that actually lens the CMB obtained an almost perfect cleaning in the absence of noise. This approach will need a careful study of possible biases coming from using the same source (the CMB) we are delensing to reconstruct the lensing effect itself. This has already been applied to data and studied. However, the fairly good efficiency of galaxies tracers might come in end to cross-check these internal biases. This will probably be needed to confirm a possible detection of gravitation waves if this relies heavily on delensing.
%=================================================================
\section{Acknowledgments}
We thank 
This work was partially supported by the Kavli Institute for Cosmological Physics at the University of Chicago through grants NSF PHY-1125897 and an endowment from the Kavli Foundation and its founder Fred Kavli.
%%%=================================================================
